---
title: "Local Explainability with e2tree: Reading the Ensemble Through the Tree"
author: "Massimo Aria, Agostino Gnasso"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Local Explainability with e2tree: Reading the Ensemble Through the Tree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
run <- requireNamespace("randomForest", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = run
)
```

## Why a local layer?

The core of **e2tree** is a *global* surrogate: a single tree built from the
ensemble's dissimilarity matrix `D`, where `D[i,j]` measures how often two
observations end up in *different* terminal nodes across the ensemble. The
tree reproduces the ensemble's partitioning geometry in a transparent form,
and `loi()` quantifies how faithfully it does so.

This vignette covers the *local* layer: a set of functions that use the
tree as a **reading lens onto the ensemble**, answering questions about a
single observation, a single node, or a single prediction. All of them are
*native* to e2tree --- they exploit the tree partition and the ensemble
proximity that defines `D`, rather than re-implementing model-agnostic tools.

| Function | Question it answers |
|---|---|
| `localLoI()` | *Where* can I trust the tree? (per-node and per-observation fidelity) |
| `eContribution()` | *Why* did this observation get this prediction? (per-instance attribution) |
| `eNeighbors()` | *Which known cases* is this observation like, in the ensemble's eyes? |
| `eHeterogeneity()` | *How decisive* is the ensemble in this region? (outcome dispersion per leaf) |
| `nodeStats()` | *Who is in this node and what makes it distinctive?* (full per-node profile, any node) |
| `plotNodeComparison()` | *How do two nodes differ?* (side-by-side region comparison) |

```{r build}
library(e2tree)
library(randomForest)

set.seed(42)
data(iris)

ensemble <- randomForest(Species ~ ., data = iris,
                         importance = TRUE, proximity = TRUE)
D <- createDisMatrix(ensemble, data = iris, label = "Species",
                     parallel = list(active = FALSE, no_cores = 1))
setting <- list(impTotal = 0.1, maxDec = 0.01, n = 2, level = 5)
tree <- e2tree(Species ~ ., iris, D, ensemble, setting)
```

## 1. `localLoI()` --- where the tree is reliable

`loi()` returns a single global number (and a within-/between-node split).
`localLoI()` disaggregates the *same* statistic to each observation and each
terminal node, so reliability becomes a map rather than a scalar. The
per-observation values average **exactly** to the global normalised LoI.

```{r localloi}
vs   <- eValidation(iris, tree, D)
prox <- proximity(vs)

ll <- localLoI(prox$ensemble, prox$e2tree, fit = tree)
print(ll)
```

The two diagnostics per node are:

- `loi_in`: within-node calibration error (how well the node's internal
  proximities match the ensemble);
- `loi_out`: ensemble proximity *lost* by separating the node's members from
  the rest (high values flag leaves that split apart observations the
  ensemble considers similar).

```{r localloi-check}
# exact decomposition of the published global index
all.equal(mean(ll$obs$loi), loi(prox$ensemble, prox$e2tree)$nloi)
```

```{r localloi-plot, fig.height = 4}
plot(ll)
```

## 2. `eContribution()` --- why this prediction

`eContribution()` gives an **exact, additive, per-instance** attribution via
the Saabas decomposition: the prediction is the root value plus the change in
node value at each split the observation passes through. Two observations that
land in different leaves receive different contributions --- and different
signs.

```{r econtribution}
ec <- eContribution(tree, newdata = iris[c(1, 84, 120), ])
print(ec)
```

```{r econtribution-additive}
# baseline + contributions reconstructs the prediction exactly
all.equal(as.numeric(ec$baseline + rowSums(ec$contributions)), ec$prediction)
```

```{r econtribution-plot, fig.height = 4}
plot(ec, obs = 2)
```

## 3. `eNeighbors()` --- the cases the ensemble finds similar

Instead of attributing to features, `eNeighbors()` explains by example: it
returns the training cases most similar to a query under the **ensemble's own
proximity** (leaf co-occurrence), together with the prototypes of the query's
leaf. Because the similarity comes from the ensemble, neighbours reflect how
the *model* groups observations, not raw feature distance --- and the
computation is identical for any supported backend.

```{r eneighbors}
nb <- eNeighbors(tree, ensemble, query = iris[c(1, 84, 120), ], k = 5)
print(nb)
```

## 4. `eHeterogeneity()` --- how decisive is the ensemble per region

`eHeterogeneity()` describes how heterogeneous the ensemble's outcomes are
*inside* each reconstructed region (terminal node). It is purely descriptive:
no prediction, no coverage guarantee. For classification it reports the regional
class mix (purity, normalised entropy, the set of non-negligible classes); for
regression it reports the centre and a central band of the regional outcomes.
A region with high entropy / `set_size > 1` (classification) or a wide band
(regression) is one where the ensemble's behaviour is intrinsically ambiguous.

```{r eheterogeneity}
h <- eHeterogeneity(tree, alpha = 0.1)
print(h)
```

This complements `localLoI()`: `localLoI()` says how *faithfully* the tree
reconstructs a region, while `eHeterogeneity()` says how *decisive* the ensemble
itself is there.

## 5. Inspecting and comparing nodes (`nodeStats`, `plotNodeComparison`)

`nodeStats()` profiles a *single* node --- terminal **or internal** --- listing
who is inside and what makes the region distinctive: per-predictor statistics
against the rest of the population (Cohen's d for numeric, Cramer's V for
categorical) and the within-region response distribution.

```{r nodestats}
term <- nodes(tree, terminal = TRUE)$node
ns <- nodeStats(tree, node = term[1])
print(ns)
plot(ns)
```

`plotNodeComparison()` puts two nodes side by side and ranks the predictors on
which they differ most --- the quickest way to answer "what separates these two
regions the ensemble built?".

```{r nodecompare, fig.height = 4}
cmp <- plotNodeComparison(tree, term[1], term[2])
head(cmp)
```

## 6. Putting it together: reliability-aware explanations

The pieces compose. Pass a `localLoI` object to `eContribution()` and every
explanation carries the fidelity of the leaf it lands in: an attribution in a
high-`loi_out` leaf should be read with caution, because the tree is a weaker
proxy for the ensemble there.

```{r together}
ec2 <- eContribution(tree, newdata = iris[c(1, 84, 120), ], reliability = ll)
ec2$reliability
```

## Regression

The same workflow applies to regression ensembles. Here `eContribution()`
decomposes a region's reconstructed value and `eHeterogeneity()` reports the
spread of outcomes within each region.

```{r regression}
data(mtcars)
set.seed(7)
tr <- sample(nrow(mtcars), 24)

ens_r  <- randomForest(mpg ~ ., data = mtcars[tr, ], ntree = 400,
                       proximity = TRUE)
D_r    <- createDisMatrix(ens_r, data = mtcars[tr, ], label = "mpg",
                          parallel = list(active = FALSE, no_cores = 1))
tree_r <- e2tree(mpg ~ ., mtcars[tr, ], D_r, ens_r,
                 list(impTotal = 0.1, maxDec = 1e-6, n = 2, level = 5))

ec_r <- eContribution(tree_r, newdata = mtcars[tr[1:3], ])
print(ec_r)
```

## One entry point: `explain()`

The functions above each answer one local question. `explain()` composes them
into a single per-instance account -- where the ensemble routed the instance,
*why* (additive attribution), which cases it is grouped with, how reliable and
how dispersed that region is -- and can attach two further layers on request.

```{r explain}
ex <- explain(tree, ensemble, newdata = iris[c(1, 84, 120), ],
              reliability = ll, counterfactual = TRUE, stability = TRUE,
              B = 50, seed = 1)
print(ex)
```

### Contrastive explanation, verified on the ensemble

`eCounterfactual()` asks what minimal feature change would move an instance into
a *different* E2Tree region. Because e2tree is interpretive, the change is only
meaningful if the **ensemble** also regroups the instance: the `validated` flag
compares the counterfactual's leaf co-occurrence proximity to the target region
against the source region. A "not validated" result is informative -- it shows
the minimal surrogate-tree change is not enough to cross the ensemble's own
grouping boundary.

```{r counterfactual}
cf <- eCounterfactual(tree, newdata = iris[c(84, 120), ], ensemble = ensemble)
print(cf)
```

### How stable is the explanation?

Most local explainers give a point estimate. `eStability()` bootstraps the
ensemble's trees to turn the explanation into a distribution, reporting a
confidence in the instance's grouping outcome and how often its nearest cases
survive resampling -- a fidelity-uncertainty statement, not a predictive
interval.

```{r stability}
st <- eStability(tree, ensemble, newdata = iris[c(1, 84, 120), ],
                 B = 100, seed = 1)
print(st)
```

## References

- Aria, M., Gnasso, A., Iorio, C., & Pandolfo, G. (2024). Explainable
  ensemble trees. *Computational Statistics*, 39(1), 3--19.
- Aria, M., Gnasso, A., Iorio, C., & Fokkema, M. (2026). Extending
  Explainable Ensemble Trees to Regression Contexts. *Applied Stochastic
  Models in Business and Industry*, 42(1), e70064.
- Saabas, A. (2014). Interpreting random forests.
- Lundberg, S.M. & Lee, S.I. (2020). From local explanations to global
  understanding with explainable AI for trees. *Nature Machine Intelligence*,
  2(1), 56--67.
