---
title: "Panel e2tree: Explaining an Ensemble on Longitudinal Data (Between / Within)"
author: "Massimo Aria, Agostino Gnasso"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Panel e2tree: Explaining an Ensemble on Longitudinal Data (Between / Within)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
run <- requireNamespace("ranger", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = run
)
```

## Panel data has two sources of variation

When the data are **panel** (or longitudinal) --- the same units observed over
time, `unit x time` in long format --- the outcome moves for two distinct
reasons:

- **between** units: some units simply sit at a higher or lower level than
  others (differences between unit averages);
- **within** units: the outcome changes over time *inside* a unit (deviations
  from that unit's own average).

These are different questions. *What makes Sweden's life expectancy higher than
Romania's?* is a between question. *What moved a given country's life expectancy
across years?* is a within question. A single explanation that mixes the two
answers neither well.

`panel_e2tree()` keeps them separate. Following the decomposition of
Mundlak (1978), it splits the **feature representation** into a between part
(unit means) and a within part (unit-demeaned deviations), grows a *separate*
[`e2tree`](e2tree-introduction.html) surrogate for each, and reconstructs the
panel additively:

$$
\hat y_{it} \;=\; \underbrace{\text{between}(\bar x_i)}_{\text{unit level}}
\;+\; \underbrace{\text{within}(x_{it} - \bar x_i)}_{\text{movement over time}} .
$$

As everywhere in **e2tree**, this is *interpretive, not predictive*: each
surrogate's quality is its **fidelity** to the ensemble it explains
($\mathrm{cor}(\text{e2tree}, \text{ensemble})^2$), and the panel figure is the
share of the *observed* outcome variance recovered by the decomposition --- a
descriptive diagnostic, never a predictive claim.

## Why a single surrogate is not enough

On panel data with high intraclass correlation (ICC) --- where the between
variance dominates --- a single *pooled* e2tree conflates the two sources: the
between variance crowds out the within signal and, with a handful of splits, the
tree cannot encode the unit levels, so fidelity collapses. Several single-model
shortcuts were tried and discarded:

| Approach | Outcome | Why it fails |
|---|---|---|
| pooled e2tree on raw features | low fidelity | conflates between / within |
| PERMANOVA decomposition of a pooled proximity | does not track the outcome ICC | proximity reflects the *feature* partition, not the outcome variance |
| single e2tree on Mundlak-augmented features `[x_bw, x_wn]` | within barely splits | between dominance crowds out within |
| **two-representation decomposition** | **high fidelity** | separates the two sources |

The lesson: the decomposition must act on the **input** (the feature
representation) and must *separate* the data, not merely augment it. That is
what `panel_e2tree()` does.

## An example panel: `panel_health`

The package ships a **simulated** country-by-year panel of life-expectancy
determinants, `panel_health` (30 synthetic countries, 2004--2019, 480 rows). It
is built so the two sources of variation are interpretable: some predictors
mostly distinguish countries (**between**: `gdp_pc`, `health_exp`, `schooling`),
while others mostly move the outcome over time within a country (**within**:
`immunization` catch-up, `undernourish` shocks). The intraclass correlation of
`life_expectancy` is high ($\approx 0.96$) --- exactly the regime where a single
pooled surrogate is overwhelmed by the between variance.

```{r build}
library(e2tree)

data(panel_health)
head(panel_health)

# intraclass correlation of the outcome (share of between variance)
mu  <- tapply(panel_health$life_expectancy, panel_health$country, mean)
icc <- var(mu[as.character(panel_health$country)]) / var(panel_health$life_expectancy)
round(icc, 2)
```

## Fitting the decomposition

The interface mirrors the rest of the package: a `formula`, the `data`, and the
name of the `unit` column. The per-representation ensembles are fitted
internally (`engine = "ranger"` by default, routed through the same adapter
layer as the rest of e2tree, so any supported backend works); a `time` column
can be passed and is reserved for future lag extensions.

```{r fit}
m <- panel_e2tree(
  life_expectancy ~ gdp_pc + health_exp + schooling +
    immunization + sanitation + undernourish,
  data = panel_health, unit = "country", time = "year",
  engine = "ranger", ntree = 400, seed = 7)
print(m)
```

The `print` method reports the three fidelities at a glance: the **between**
surrogate and the **within** surrogate against their respective ensembles, and
the **panel** reconstruction against the observed outcome. Note which variables
each tree splits on --- the between tree picks up the cross-country level drivers
(`gdp_pc`, `health_exp`), while the within tree picks up what moves a country
over time, led by `immunization`.

## Reading the decomposition

`summary()` expands the report: the panel shape (balanced or not), the
fidelities, the **decomposed variable importance** (between vs within), and any
features dropped from a representation. Dropping is automatic and meaningful: a
time-invariant feature becomes constant after within-demeaning and is removed
from the *within* tree, while a feature with no between-unit variation is removed
from the *between* tree. (In `panel_health` every predictor carries both kinds
of variation, so none is dropped.)

```{r summary}
summary(m)
```

The decomposed importance is the payoff: it tells you *separately* which
features explain where countries stand (income and health spending, between) and
which explain how a country moves over time (immunization gains, undernourishment
shocks, within) --- a distinction a pooled importance ranking cannot draw.

## The two trees

Each component is an ordinary `e2tree`, so it plots with the usual
`plot_e2tree()`. `plot.e2panel()` draws either or both.

```{r plot-between, fig.height = 4.5}
plot(m, component = "between")
```

```{r plot-within, fig.height = 4.5}
plot(m, component = "within")
```

## The necessity argument, empirically

Does the decomposition actually buy fidelity over a single pooled surrogate? We
fit one e2tree on the raw features and reconstruct the observed outcome, then
compare.

```{r necessity}
f <- life_expectancy ~ gdp_pc + health_exp + schooling +
  immunization + sanitation + undernourish

ens_pool  <- ranger::ranger(f, data = panel_health,
                            num.trees = 400, importance = "permutation")
D_pool    <- createDisMatrix(ens_pool, data = panel_health, label = "life_expectancy",
                             parallel = list(active = FALSE, no_cores = 1))
tree_pool <- e2tree(f, panel_health, D_pool, ens_pool,
                    setting = list(impTotal = 0.1, maxDec = 1e-6, n = 2, level = 5))

pooled_fit      <- as.numeric(predict(tree_pool, newdata = panel_health)$fit)
pooled_fidelity <- cor(pooled_fit, panel_health$life_expectancy)^2

c(pooled = pooled_fidelity, decomposed_panel = m$fidelity_panel)
```

The decomposed reconstruction matches or exceeds the pooled surrogate. The same
pattern holds, more sharply, in the published life-expectancy application
(137--144 real countries, 2000--2024, ICC $\approx 0.86$): pooled fidelity
$\approx 0.39$ versus a decomposed panel fidelity of $0.82$--$0.83$ (between
$0.93$, within $0.66$ with the full feature set). The higher the ICC, the more a
single surrogate is overwhelmed by the between variance, and the more the
decomposition is needed.

## Prediction on new observations

`predict()` reconstructs the additive explanation for new rows: it maps the
between prediction by unit and runs the within tree on the deviation from the
*training* unit mean. The output separates the two contributions.

```{r predict}
newrows <- panel_health[c(1, 17, 33), ]   # one row from each of three countries
predict(m, newdata = newrows)
```

For a unit unseen in training the between prediction falls back to the mean over
training units and the within contribution is set to zero (with a warning),
since the unit's reference level is unknown.

## Diagnostics and scope

**Within signal.** The within tree can only be as informative as the *within
variation* of the features. With few time-varying predictors it degenerates;
`panel_e2tree()` reports the mean within-variance share of the features
(`summary()`) and warns when it is low. In the life-expectancy study, moving
from 8 to 11 features --- adding the ILR coordinates of the financing mix ---
lifted the within fidelity from $\approx 0.01$ to $0.66$.

**Unbalanced panels.** Missing rows are handled (complete cases per the selected
columns; unit means computed with `na.rm = TRUE`); `summary()` flags whether the
panel is balanced.

**Regression only (for now).** This first version targets numeric outcomes with
numeric predictors. The between component generalises naturally to
classification (unit proportions), but a "within deviation" of a categorical
target is an open methodological direction; a factor outcome raises an explicit
error. Future work also includes within trees with *lags* and native
between/within significance tests.

## References

- Mundlak, Y. (1978). On the pooling of time series and cross section data.
  *Econometrica*, 46(1), 69--85.
- Aria, M., Gnasso, A., Iorio, C., & Pandolfo, G. (2024). Explainable
  ensemble trees. *Computational Statistics*, 39(1), 3--19.
- Aria, M., Gnasso, A., Iorio, C., & Fokkema, M. (2026). Extending
  Explainable Ensemble Trees to Regression Contexts. *Applied Stochastic
  Models in Business and Industry*, 42(1), e70064.
