---
title: "Using e2tree with XGBoost, GBM, LightGBM, and CatBoost"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using e2tree with XGBoost, GBM, LightGBM, and CatBoost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # examples require optional packages; set eval=TRUE locally
)
```

## Overview

Starting from version 1.1.0, `e2tree` supports the following tree ensemble
backends in addition to `randomForest` and `ranger`:

| Package | Model class | Task |
|---------|-------------|------|
| **xgboost** | `xgb.Booster` | classification, regression |
| **gbm** | `gbm` | classification, regression |
| **lightgbm** | `lgb.Booster` | classification, regression |
| **catboost** | `catboost.CatBoost` / `catboost.Model` | classification, regression |

The workflow is identical regardless of the backend: train a model, build the
dissimilarity matrix with `createDisMatrix()`, then call `e2tree()`.

---

## XGBoost

### Classification (iris)

```{r xgb-clf}
library(e2tree)
library(xgboost)

data(iris)
set.seed(42)
n  <- floor(0.75 * nrow(iris))
tr <- iris[sample(nrow(iris), n), ]
va <- iris[setdiff(seq_len(nrow(iris)), as.integer(rownames(tr))), ]

# XGBoost requires a numeric matrix and 0-indexed integer labels
X_tr <- as.matrix(tr[, 1:4])
y_tr <- as.integer(tr$Species) - 1L
dm_tr <- xgb.DMatrix(data = X_tr, label = y_tr)

ensemble <- xgb.train(
  params  = list(objective  = "multi:softmax",
                 num_class  = 3,
                 max_depth  = 4,
                 eta        = 0.3),
  data    = dm_tr,
  nrounds = 100,
  verbose = 0
)

# Attach the response back to the data.frame so the formula in e2tree()
# can find it; createDisMatrix() will use it to annotate the dissimilarity
# matrix (in classification, `label` is optional but recommended).
tr_xgb         <- tr[, 1:4]
tr_xgb$Species <- tr$Species

D <- createDisMatrix(ensemble, data = tr_xgb, label = "Species",
                     parallel = list(active = FALSE, no_cores = 1))

setting  <- list(impTotal = 0.1, maxDec = 0.01, n = 2, level = 5)
tree_xgb <- e2tree(Species ~ ., data = tr_xgb, D = D,
                   ensemble = ensemble, setting = setting)
print(tree_xgb)
```

### Regression (mtcars)

For regression backends, `createDisMatrix()` needs the response column to
compute the dissimilarity scale. Pass the full data frame (predictors plus
response) and the name of the response column via the `label` argument.

```{r xgb-reg}
library(xgboost)

data(mtcars)
set.seed(42)
n  <- floor(0.75 * nrow(mtcars))
tr <- mtcars[sample(nrow(mtcars), n), ]

X_tr  <- as.matrix(tr[, -1])
y_tr  <- tr$mpg
dm_tr <- xgb.DMatrix(data = X_tr, label = y_tr)

ensemble <- xgb.train(
  params  = list(objective = "reg:squarederror", max_depth = 4, eta = 0.3),
  data    = dm_tr,
  nrounds = 100,
  verbose = 0
)

# `data = tr` carries the response column too; the XGBoost adapter
# automatically trims the matrix to the features used at training time.
D    <- createDisMatrix(ensemble, data = tr, label = "mpg",
                        parallel = list(active = FALSE, no_cores = 1))
tree <- e2tree(mpg ~ ., data = tr, D = D, ensemble = ensemble,
               setting = list(impTotal = 0.1, maxDec = 1e-6, n = 2, level = 5))
print(tree)
```

---

## GBM

### Classification (iris, binary)

`gbm` expects a 0/1 numeric response for the `bernoulli` distribution, while
`e2tree` expects a factor response for classification. We therefore train
`gbm` on the integer column and pass a factor copy of the same column to
`e2tree`.

```{r gbm-clf}
library(gbm)

data(iris)
set.seed(42)
df       <- iris
df$is_setosa     <- as.integer(df$Species == "setosa")
df$is_setosa_fct <- factor(df$is_setosa, levels = c(0L, 1L))
n  <- floor(0.75 * nrow(df))
tr <- df[sample(nrow(df), n), ]

ensemble <- gbm(is_setosa ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
                data              = tr,
                distribution      = "bernoulli",
                n.trees           = 200,
                interaction.depth = 4,
                verbose           = FALSE)

D    <- createDisMatrix(ensemble,
                        data     = tr[, c("Sepal.Length","Sepal.Width",
                                          "Petal.Length","Petal.Width",
                                          "is_setosa_fct")],
                        label    = "is_setosa_fct",
                        parallel = list(active = FALSE, no_cores = 1))
tree <- e2tree(is_setosa_fct ~ Sepal.Length + Sepal.Width +
                               Petal.Length + Petal.Width,
               data = tr, D = D, ensemble = ensemble,
               setting = list(impTotal = 0.1, maxDec = 0.01, n = 2, level = 5))
print(tree)
```

### Regression (mtcars)

`gbm` requires `nTrain * bag.fraction > 2 * n.minobsinnode + 1`, which fails
on small training sets such as 24-row `mtcars` with the default settings.
Lower `n.minobsinnode` and raise `bag.fraction` to keep the example
self-contained.

```{r gbm-reg}
library(gbm)

data(mtcars)
set.seed(42)
n  <- floor(0.75 * nrow(mtcars))
tr <- mtcars[sample(nrow(mtcars), n), ]

ensemble <- gbm(mpg ~ ., data = tr,
                distribution      = "gaussian",
                n.trees           = 200,
                interaction.depth = 4,
                n.minobsinnode    = 2,
                bag.fraction      = 0.8,
                verbose           = FALSE)

D    <- createDisMatrix(ensemble, data = tr, label = "mpg",
                        parallel = list(active = FALSE, no_cores = 1))
tree <- e2tree(mpg ~ ., data = tr, D = D, ensemble = ensemble,
               setting = list(impTotal = 0.1, maxDec = 1e-6, n = 2, level = 5))
print(tree)
```

---

## LightGBM

### Classification (iris)

```{r lgb-clf}
library(lightgbm)

data(iris)
set.seed(42)
n  <- floor(0.75 * nrow(iris))
tr <- iris[sample(nrow(iris), n), ]

X_tr <- as.matrix(tr[, 1:4])
y_tr <- as.integer(tr$Species) - 1L
ds   <- lgb.Dataset(X_tr, label = y_tr)

ensemble <- lgb.train(
  params  = list(objective  = "multiclass",
                 num_class  = 3,
                 num_leaves = 15,
                 verbose    = -1),
  data    = ds,
  nrounds = 100
)

tr_lgb         <- tr[, 1:4]
tr_lgb$Species <- tr$Species

D <- createDisMatrix(ensemble, data = tr_lgb, label = "Species",
                     parallel = list(active = FALSE, no_cores = 1))

tree <- e2tree(Species ~ ., data = tr_lgb, D = D, ensemble = ensemble,
               setting = list(impTotal = 0.1, maxDec = 0.01, n = 2, level = 5))
print(tree)
```

### Regression (mtcars)

```{r lgb-reg}
library(lightgbm)

data(mtcars)
set.seed(42)
n  <- floor(0.75 * nrow(mtcars))
tr <- mtcars[sample(nrow(mtcars), n), ]

X_tr <- as.matrix(tr[, -1])
y_tr <- tr$mpg
ds   <- lgb.Dataset(X_tr, label = y_tr)

ensemble <- lgb.train(
  params  = list(objective        = "regression",
                 num_leaves       = 8,
                 min_data_in_leaf = 2,
                 learning_rate    = 0.1,
                 verbose          = -1),
  data    = ds,
  nrounds = 200
)

# Pass the response column to createDisMatrix() via `label`. The
# LightGBM adapter selects the columns it needs through the booster's
# stored feature names, so any extra columns in `data` are ignored.
D    <- createDisMatrix(ensemble, data = tr, label = "mpg",
                        parallel = list(active = FALSE, no_cores = 1))
tree <- e2tree(mpg ~ ., data = tr, D = D, ensemble = ensemble,
               setting = list(impTotal = 0.1, maxDec = 1e-6, n = 2, level = 5))
print(tree)
```

---

## CatBoost

CatBoost is distributed separately and must be installed from GitHub:

```{r catboost-install, eval = FALSE}
# install.packages("remotes")
remotes::install_github(                                                                                                        
    "catboost/catboost",                 
    subdir = "catboost/R-package",       
    INSTALL_opts = c("--no-staged-install", "--no-multiarch", "--no-test-load")
  ) 
```

### Classification (iris)

```{r catboost-clf}
library(catboost)

data(iris)
set.seed(42)
n  <- floor(0.75 * nrow(iris))
tr <- iris[sample(nrow(iris), n), ]

pool_tr <- catboost.load_pool(
  data  = tr[, 1:4],
  label = as.integer(tr$Species) - 1L
)

ensemble <- catboost.train(
  pool_tr,
  params = list(loss_function = "MultiClass",
                iterations    = 100,
                logging_level = "Silent")
)
# Store label column name so the adapter can strip it from newdata
attr(ensemble, "e2tree_label") <- "Species"

D    <- createDisMatrix(ensemble, data = tr, label = "Species",
                        parallel = list(active = FALSE, no_cores = 1))
tree <- e2tree(Species ~ ., data = tr, D = D, ensemble = ensemble,
               setting = list(impTotal = 0.1, maxDec = 0.01, n = 2, level = 5))
print(tree)
```

### Regression (mtcars)

```{r catboost-reg}
library(catboost)

data(mtcars)
set.seed(42)
n  <- floor(0.75 * nrow(mtcars))
tr <- mtcars[sample(nrow(mtcars), n), ]

pool_tr <- catboost.load_pool(
  data  = tr[, -1],
  label = tr$mpg
)

ensemble <- catboost.train(
  pool_tr,
  params = list(loss_function  = "RMSE",
                iterations     = 100,
                logging_level  = "Silent")
)
attr(ensemble, "e2tree_label") <- "mpg"

D    <- createDisMatrix(ensemble, data = tr, label = "mpg",
                        parallel = list(active = FALSE, no_cores = 1))
tree <- e2tree(mpg ~ ., data = tr, D = D, ensemble = ensemble,
               setting = list(impTotal = 0.1, maxDec = 1e-6, n = 2, level = 5))
print(tree)
```

---

## Adding a new backend

To support a further model class `MyEnsemble`, implement three S3 methods and
register them in `NAMESPACE`:

```r
# In R/adapters.R (or a separate R/adapter_mymodel.R)

get_ensemble_type.MyEnsemble <- function(ensemble) {
  # return "classification" or "regression"
}

extract_terminal_nodes.MyEnsemble <- function(ensemble, data) {
  # return data.frame of (n_obs × n_trees) terminal node IDs
}

get_ensemble_predictions.MyEnsemble <- function(ensemble, data, type) {
  # return numeric vector of length n_obs
}
```

No changes to `createDisMatrix()`, `e2tree()`, or any other core function are
required.
